---
title: "CS01: Biomarkers of Recent Use"
author: "Yuancheng Cao, Vinuthna Hasthi, Darwin Yu, Nina Hong"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
# control global Rmd chunk settings
knitr::opts_chunk$set(warning=FALSE)
```

### Load packages

```{r load-packages, message=FALSE}
library(tidyverse)
library(ggplot2)
```

## Introduction

In this case study, we seek to determine the best bio-markers of recent THC use within a 3-hour period by evaluating compounds across three matrices: blood, oral fluid, and breath. Additionally, we aim to determine the optimal THC concentration cutoffs for each matrix within this timeframe and to examine how treatment groups may influence THC levels. This analysis is important given the legal implications of THC detection in impaired driving cases. Nineteen states have implemented per se or zero-tolerance laws for THC, with per se limits varying between 1-5 ng/mL THC in blood [^1]. These limits suggest significant regulatory interest in identifying reliable biological markers of recent cannabis use, particularly in roadside testing contexts.

The study’s data set includes three matrices–blood, oral fluid, and breath–with a total of 573 participants across them. Blood and oral fluid matrices allow detection of multiple THC-related compounds (8 in blood, 7 in oral fluid), while breath analysis focuses solely on one compound. Blood and oral fluid matrices have previously demonstrated relatively stable THC levels over time, providing a basis for exploring how they reflect THC concentrations within a 3-hour window [^2]. Breath testing, however, remains a newer field of investigation, and its reliability for THC detection within a short timeframe is still under review. Identifying the optimal biological matrix for THC detection in recent use is important for law enforcement to establish reliable, science-based standards for roadside impairment testing.

The analysis also seeks to address the legal challenges posed by varying state regulations. For example, states like Colorado utilize a “reasonable inference” approach, where blood concentrations above 5 ng/ml THC suggest impairment [^3]. These differing legal standards show the importance of identifying clear, evidence-based cutoff values for THC detection across different matrices. In this case study, we aim to determine which matrix is most effective for detecting recent THC use within a 3-hour window and to identify the optimal cutoff values for each matrix. This analysis will help us understand how different matrices perform in detecting THC levels over time and provide insights into their potential effectiveness in practical settings.

## Questions

1. Which matrix–blood, oral fluid, or breath–is most effective for detecting recent THC use within a 3-hour window?
2. What is the optimal cutoff within this 3-hour timeframe for detecting THC in each matrix?
3. How does the treatment group (Placebo, 5.9% THC, or 13.4% THC) affect the detection of recent THC use in each matrix?


## The Data

The data set that we will be using for our case study is provided to us by Professor Ellis in our data/ folder. The data provided will be in the form of csv files.

### Data Import

In order to begin our analysis, we imported our raw data of whole blood, breath, and oral fluids from our data/ folder.

```{r, message=FALSE}
WB <- read_csv("data/Blood.csv")
BR <- read_csv("data/Breath.csv")
OF <- read_csv("data/OF.csv")
```
### Data Wrangling

Now that we have our data properly imported, we need to wrangle our data in R before starting our analysis. 

To begin our data wrangling, we imported the data as tables and modified the columns, transforming the data to improve its structure. There are three data set tables for each fluid type, blood, oral, and breath, in which we reformatted the entries in the `Treatment` column for better readability. We also converted all column names to snake case to simplify future coding. We renamed the columns for all compounds in each data set to ensure clear differentiation. Note that not every fluid type contains all of the compounds. To create time intervals, we turned the `time_from_start`column into hour intervals in `timepoint`.

We start with the Whole Blood table.

```{r, message=FALSE}
WB <- WB |> 
  mutate(Treatment = fct_recode(Treatment, 
                                "5.9% THC (low dose)" = "5.90%",
                                "13.4% THC (high dose)" = "13.40%"),
         Treatment = fct_relevel(Treatment, "Placebo", "5.9% THC (low dose)")) |> 
  janitor::clean_names() |>
  rename(thcoh = x11_oh_thc,
         thccooh = thc_cooh,
         thccooh_gluc = thc_cooh_gluc,
         thcv = thc_v) |>
  mutate(timepoint = case_when(time_from_start < 0 ~ "pre-smoking",
                               time_from_start > 0 & time_from_start <= 60 ~ "0-60 min",
                               time_from_start > 61 & time_from_start <= 120 ~ "0-120 min",
                               time_from_start > 121 & time_from_start <= 180 ~ "0-180 min",
                               time_from_start > 180 ~ "181+ min"))
```

Then, the Oral Fluid table.

```{r, message=FALSE}
OF <- OF |> 
  mutate(Treatment = fct_recode(Treatment, 
                                "5.9% THC (low dose)" = "5.90%",
                                "13.4% THC (high dose)" = "13.40%"),
         Treatment = fct_relevel(Treatment, "Placebo", "5.9% THC (low dose)")) |> 
  janitor::clean_names() |>
  rename(thcoh = x11_oh_thc,
         thcv = thc_v,
         fluid_type = fluid) |>
  mutate(timepoint = case_when(time_from_start < 0 ~ "pre-smoking",
                               time_from_start > 0 & time_from_start <= 60 ~ "0-60 min",
                               time_from_start > 61 & time_from_start <= 120 ~ "0-120 min",
                               time_from_start > 121 & time_from_start <= 180 ~ "0-180 min",
                               time_from_start > 180 ~ "181+ min"))
```

Last, but not least, the Breath table.

```{r, message=FALSE}
BR <- BR |> 
  mutate(Treatment = fct_recode(Treatment, 
                                "5.9% THC (low dose)" = "5.90%",
                                "13.4% THC (high dose)" = "13.40%"),
         Treatment = fct_relevel(Treatment, "Placebo", "5.9% THC (low dose)")) |> 
  janitor::clean_names() |>
  rename(thc = thc_pg_pad,
         fluid_type = fluid) |>
  mutate(timepoint = case_when(time_from_start < 0 ~ "pre-smoking",
                               time_from_start > 0 & time_from_start <= 60 ~ "0-60 min",
                               time_from_start > 61 & time_from_start <= 120 ~ "0-120 min",
                               time_from_start > 121 & time_from_start <= 180 ~ "0-180 min",
                               time_from_start > 180 ~ "181+ min"))
```

After cleaning our data, we combined the three dataset tables into a single data set and stored in a new csv.

```{r, message=FALSE}
combined <- bind_rows(list(WB, OF, BR))
write_csv(combined, "data/combined_data.csv")
```

## Analysis

```{r compounds, message=FALSE}
combined_long <- combined |> 
  select(1:5,time_from_start,everything()) |>
  pivot_longer(7:15)

invisible(ggplot(combined_long, aes(x = value)) + 
  geom_histogram() +
  facet_wrap(~name))

invisible(combined_long |>
  mutate(group_compound = paste0(fluid_type, ": ", name)) |>
  ggplot(aes(x = value)) + 
  geom_histogram() + 
  facet_wrap(~group_compound, scales = "free"))

invisible(combined_long |> 
  filter(name == "thc") |>
  ggplot(aes(x = group, y = value)) + 
  geom_boxplot() +
  facet_wrap(~fluid_type, scales = "free"))

invisible(combined_long |> 
  filter(name == "thc") |>
  ggplot(aes(x = treatment, y = value)) + 
  geom_boxplot() +
  facet_wrap(~fluid_type, scales = "free"))

invisible(combined_long |> 
  filter(name == "thc", (timepoint == "0-30 min" | timepoint == "0-40 min")) |>
  ggplot(aes(x = treatment, y = value)) + 
  geom_boxplot() +
  facet_wrap(~fluid_type, scales = "free"))

plot_scatter_time <- function(matrix) {
  combined_long |> 
      filter(!is.na(time_from_start), fluid_type==matrix) |>
      ggplot(aes(x=time_from_start, y=value, color=group)) + 
        geom_point() +
        facet_wrap(~name, scales="free") +
        scale_color_manual(values=c("#19831C", "#A27FC9")) +
        theme_classic() +
        labs(title=paste("Compound Levels Across Time for", matrix),
             x="Time From Start (min)",
             y="Concentration (pg/mL)") +
        theme(legend.position="bottom",
              legend.title=element_blank(),
              strip.background=element_blank(),
              plot.title.position="plot") 
}

plot_scatter_time(matrix = "WB")
plot_scatter_time(matrix = "OF")
plot_scatter_time(matrix = "BR")
```
From each compound, we have

```{r thc across fluids, message=FALSE}

ggplot(combined, aes(x = time_from_start, y = thc)) +
  geom_point(size = 1, color = "lightgreen") +
  facet_wrap(~ fluid_type, scales = "free_y") +

  labs(
    title = "THC Levels across Time in each Fluid Type",
    x = "Time from Start (minutes)",
    y = "Concentration (pg/mL)",
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "top"
  )
```


```{r WB all dosages, message=FALSE}
ggplot(combined, aes(x = timepoint, y = thc, color = treatment, group = treatment)) +
  geom_line() +
  geom_point(position = position_jitter(width = 0.2), size = 2) +  # Add jitter for better visibility
  facet_wrap(~ fluid_type, scales = "free_y") +  # Create separate plots for each fluid type
  labs(
    title = "THC Levels Across Fluid Types Over Time",
    x = "Time Interval",
    y = "THC Level",
    color = "Dosage"
  ) +
  theme_minimal(base_size = 14) +  # Increase base font size for readability
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "top"
  )
```


```{r WB of the first 3 hours, message=FALSE}
ggplot(combined %>% filter(fluid_type == "WB" & time_from_start >= 0 & time_from_start <= 180), aes(x = time_from_start, y = thc, color = treatment, group = treatment)) +
  geom_line() +
  geom_point(position = position_jitter(width = 0.2), size = 2) +  # Add jitter for better visibility
  labs(
    title = "THC Levels Across Water Blood Over Time",
    x = "Time Interval",
    y = "THC Level",
    color = "Dosage"
  ) +
  theme_minimal(base_size = 14) +  # Increase base font size for better readability
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    legend.position = "bottom"
  )
```

```{r thc in wb, message=FALSE}
ggplot(combined %>% filter(fluid_type == "WB" & time_from_start >= 0 & time_from_start <= 60), aes(x = time_from_start, y = thc, color = treatment, group = treatment)) +
  geom_line() +
  geom_point(position = position_jitter(width = 0.2), size = 2) +  # Add jitter for better visibility
  labs(
    title = "THC Levels Across Water Blood Over Time",
    x = "Time Interval",
    y = "THC Level",
    color = "Dosage"
  ) +
  theme_minimal(base_size = 14) +  # Increase base font size for better readability
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    legend.position = "bottom"
  )
```


```{r}
library(patchwork)
```


```{r thc in wb 2, message=FALSE}
ggplot(combined %>% filter(fluid_type == "WB" & time_from_start >= 0 & time_from_start <= 60), aes(x = time_from_start, y = thc, color = treatment, group = treatment)) +
  geom_line() +
  geom_point(position = position_jitter(width = 0.2), size = 2) +
  facet_wrap(~treatment) +
  labs(
    title = "THC Levels Across Water Blood Over Time",
    x = "Time Interval",
    y = "THC Level",
    color = "Dosage"
  ) +
  theme_minimal(base_size = 14) +  # Increase base font size for better readability
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    legend.position = "bottom"
  )
```
```{r thc in of, message=FALSE}
ggplot(combined %>% filter(fluid_type == "OF" & time_from_start >= 0 & time_from_start <= 60), aes(x = time_from_start, y = thc, color = treatment, group = treatment)) +
  geom_line() +
  geom_point(position = position_jitter(width = 0.2), size = 2) +
  facet_wrap(~treatment) +
  labs(
    title = "THC Levels Across Oral Fluid Over Time",
    x = "Time Interval",
    y = "THC Level",
    color = "Dosage"
  ) +
  theme_minimal(base_size = 14) +  # Increase base font size for better readability
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    legend.position = "bottom"
  )
```

Specificity and Sensitivity

As we focus on THC as the main compound to our study for recent THC use, it would be best to explore the [bold] specificity [bold] pf THC as we want to be the most certain that using knowledge that they are either a frequent or occasional user, are within the time interval of one hour, and tested through WB, we are more likely to identify with high specificity if the individual has used in the past hour (less false positives).

With respect to a study that found the highest accuracy of yielding 96% for a "molar metabolite ratio 1", the study had measured whole blood THC and metabolites in the time interval of 30 minutes "after starting a 15-minute interval" of cannabis intake, where they had found with THC alone the cut-point yielded "88% specificity, 73% sensitivity, and 80% accuracy" for identifying recent cannabis use. 

Within our study, we want to utilize the focus for specificity as we calculate our data to include the cutoff of similarly to their cut-point of 5.3 mu grams/L, which is 5.3 ng/mL. In the study, it has yielded around 75% specificity and sensitivity, so this will be our baseline with the data we have currently. As the data displays THC concentration of 

Change Combined Data to include only "Frequent user" and "Occassional user" instead of "experienced" and "non-experienced"
```{r}
combined_user_long <- combined_long |>
  mutate(group = fct_recode(group, "Frequent user" = "Experienced user", "Occasional user" = "Not experienced user"))
```

Range, Median, Mean of THC in Whole Blood\
```{r}
wbthc_range <- range(WB$thc)
wbthc_median <- median(WB$thc, na.rm = TRUE)
wbthc_mean <- mean(WB$thc, na.rm = TRUE)

WBTHC_summary <- list(
  range = wbthc_range,
  median = wbthc_median,
  wbthc_mean
)

print(WBTHC_summary)
```

Calculation Function for Specificity and Sensitivity
```{r sens and spec}
make_calculations <- function(dataset, cutoff, compound, timepoint_use){
  ## remove NAs
  df <- dataset |>
    select(treatment, {{ compound }}, timepoint) |>
    filter(timepoint == timepoint_use, !is.na({{ compound }}))

  if(nrow(df)>0){
    if(timepoint_use == "pre-smoking"){
      output <- df |> 
        summarize(TP = 0,
                  FN = 0,
                  FP = sum(!!sym(compound) >= cutoff),
                  TN = sum(!!sym(compound) < cutoff)) 
    }else{
      if(cutoff == 0){
        output_pre <- df |> 
          filter(timepoint_use == "pre-smoking") |>
          summarize(TP = 0,
                    FN = 0,
                    FP = sum(!!sym(compound) >= cutoff),
                    TN = sum(!!sym(compound) < cutoff)) 
        
        output <- df |> 
          filter(timepoint_use != "pre-smoking") |>
          summarize(TP = sum(treatment != "Placebo" & !!sym(compound) > cutoff),
                    FN = sum(treatment != "Placebo" & !!sym(compound) <= cutoff),
                    FP = sum(treatment == "Placebo" & !!sym(compound) > cutoff),
                    TN = sum(treatment == "Placebo" & !!sym(compound) < cutoff))
        
        output <- output + output_pre
      }else{
        output_pre <- df |> 
          filter(timepoint_use == "pre-smoking") |>
          summarise(TP = 0,
                    FN = 0,
                    FP = sum(!!sym(compound) >= cutoff),
                    TN = sum(!!sym(compound) < cutoff)) 
        
        output <- df |> 
          filter(timepoint_use != "pre-smoking") |>
          summarise(TP = sum(treatment != "Placebo" & !!sym(compound) >= cutoff),
                    FN = sum(treatment != "Placebo" & !!sym(compound) < cutoff),
                    FP = sum(treatment == "Placebo" & !!sym(compound) >= cutoff),
                    TN = sum(treatment == "Placebo" & !!sym(compound) < cutoff))
        
        output <- output + output_pre
      }
    }
  
  # clean things up; make calculations on above values
  output <- output |>
    mutate(detection_limit = cutoff,
           compound = compound,
           time_window = timepoint_use,
           NAs = nrow(dataset) - nrow(df),
           N = nrow(dataset),
           Sensitivity = (TP/(TP + FN)), 
           Specificity = (TN /(TN + FP)),
           PPV = (TP/(TP+FP)),
           NPV = (TN/(TN + FN)),
           Efficiency = ((TP + TN)/(TP + TN + FP + FN))*100
    )
  return(output)
}
}
```

Map for each matrix
```{r}
sens_spec_cpd <- function(dataset, cpd, timepoints){
  args2 <- list(start = timepoints$start, 
                stop = timepoints$stop, 
                tpt_use = timepoints$timepoint)
  out <- args2 |> 
    pmap_dfr(make_calculations, dataset, compound = cpd)
  return(out)
}
```

Choosing the cutoffs for THC, we used 5 ng/mL as our baseline representation. 
```{r}
# specify which calculations to make
cutoffs <- c(0.5, 1, 5, 7.5, 10)
compounds <- combined_user_long |> filter(fluid_type=="WB") |> filter(!is.na(value)) |> distinct(name) |> pull(name)
WB_timepoints <- c("pre-smoking", "0-60 min", "0-120 min", "181+ min")
WB <- combined |> filter(fluid_type=="WB")

# Specify all parameter combinations
param_grid <- expand.grid(
  cutoffs = cutoffs,
  compounds = compounds,
  timepoint_use = WB_timepoints)


# Calculate for all cutoff-compound-timepoint combinations
WB_ss <- purrr::pmap_dfr(param_grid, ~ make_calculations(dataset=WB, cutoff = ..1, compound = as.character(..2), timepoint_use = ..3))

WB_ss
```

Sensitivity and Specificity Plot
```{r}
plot_cutoffs <- function(dataset, timepoint_use_variable, tissue, cpd){
    # control colors and lines used in plots
    col_val = c("#D9D9D9", "#BDBDBD", "#969696", "#636363", "#252525")
    lines = rep("solid", 5)
    
    # prep data
    df_ss <- dataset |> 
      filter(compound == cpd) |>
      mutate(time_window = fct_relevel(as.factor(time_window), levels(timepoint_use_variable)),
             detection_limit = as.factor(detection_limit),
             Sensitivity =  round(Sensitivity*100, 0),
             Specificity =  round(Specificity*100, 0))        
      
    # plot sensitivity
    p1 <- df_ss |> 
      ggplot(aes(x = time_window, y = Sensitivity, 
                 color = detection_limit)) + 
      geom_line(linewidth = 1.2, aes(group = detection_limit, 
                                linetype = detection_limit)) + 
      geom_point(show.legend=FALSE) + 
      ylim(0,100) +
      scale_x_discrete(labels = function(x) str_wrap(x, width = 5)) +
      scale_linetype_manual(values=lines) +
      scale_color_manual(values = col_val, name = "Cutoff \n (ng/mL)",
                        guide = guide_legend(override.aes = list(linetype = c(1),
                        shape = rep(NA, length(lines))) )) +
      theme_classic() +
      theme(plot.title.position = "plot",
            axis.title = element_text(size=14),
            axis.text = element_text(size=10),
            legend.position = "none", 
            panel.grid = element_blank(),
            strip.background = element_blank()
            ) +
      guides(linetype = "none") +
      labs(x = "Time Window (min)", 
           y = "Sensitivity", 
           title = paste0(tissue,": ", toupper(cpd)) )
  
  # plot specificity
  p2 <- df_ss |> 
      ggplot(aes(x = time_window, y = Specificity,
                 group = detection_limit, 
                 color = detection_limit, 
                 linetype = detection_limit)) + 
      geom_line(linewidth = 1.2) +
      geom_point() + 
      ylim(0,100) +
      scale_color_manual(values = col_val) +
      scale_x_discrete(labels = function(x) str_wrap(x, width = 5)) +
      scale_linetype_manual(values = lines, 
                            guide = guide_legend(override.aes = list(linetype = "solid",
                                                                     shape = rep(NA, length(lines))) )) +
      theme_classic() +
      theme(axis.title = element_text(size=14),
            axis.text = element_text(size=10),
            legend.position = c(0.35, 0.25),
            panel.grid = element_blank(),
            strip.background = element_blank()) +
      labs(x = "Time Window", 
           y = "Specificity",
           title = "" )
  
  # combine plots (uses patchwork)
  p1 + p2
}
```

Whole Blood Content for THC Cutoff Plot Over Time
```{r}
plot_cutoffs(dataset=WB_ss, 
             timepoint_use_variable=WB$timepoint, 
             tissue="Blood", 
             cpd="thc")
```

We can see that with our plot of specificity, 5 ng/mL and more of THC in blood yield over 90% in specificity and over  73% sensitivity to identify individuals who do not have any THC in the system, indicating that similar to prior research, there is validity to the concentration of 5.3 ng/mL for THC to be likely to be identifiable of THC and can provide decisions on when to exclude false positives of THC in system. With sensitivity, although high in the first hour, the time periods drop directly after to lower than 25% sensitvity after 3 hours, so we want to note for future concentrations that have an increase dose of 5 ng/mL may opt for higher sensitivity. 

Now, to compare with other matrix's to see if THC has a higher sensitivity and specificity compared to Oral Fluid and Breath. 

### Oral Fluid
```{r}
# specify which calculations to make
cutoffs <- c(0.5, 1, 5, 7.5, 10)
compounds <- combined_user_long |> filter(fluid_type=="OF") |> filter(!is.na(value)) |> distinct(name) |> pull(name)
OF_timepoints <- c("pre-smoking","0-60 min","0-120 min",
                   "181+ min") 
OF <- combined |> filter(fluid_type=="OF")

# Specify all parameter combinations
param_grid <- expand.grid(
  cutoffs = cutoffs,
  compounds = compounds,
  timepoint_use = OF_timepoints)

# Calculate for all cutoff-compound-timepoint combinations
OF_ss <- purrr::pmap_dfr(param_grid, ~ make_calculations(dataset=OF, cutoff = ..1, compound = as.character(..2), timepoint_use = ..3))

OF_ss
```
```{r}
plot_cutoffs(dataset=OF_ss, 
             timepoint_use_variable=OF$timepoint, 
             tissue="Oral Fluid", 
             cpd="thc")
```
When looking at Oral Fluid, we can see that for 5 ng/mL, the trend of high sensitivity and specificity continues as for the presmoking period in specificity and for the first hour for both all yield high percentages, indicating that oral fluid also contributes to having a high likelihood to detect THC even in small amounts. However, although sensitivity is higher for all concentrations in the first hour, there is a large falloff in the first hour, distinctly in detection of false positives in specificity for lower doses within the hour interval, which does not yield more accurate results compared to whole blood.

### Breath Level
```{r}
# specify which calculations to make
cutoffs <- c(0.5, 1, 5, 7.5, 10)
compounds <- combined_user_long |> filter(fluid_type=="BR") |> filter(!is.na(value)) |> distinct(name) |> pull(name)
BR_timepoints <- c("pre-smoking","0-60 min","0-120 min",
                   "181+ min") 
BR <- combined |> filter(fluid_type=="BR")

# Specify all parameter combinations
param_grid <- expand.grid(
  cutoffs = cutoffs,
  compounds = compounds,
  timepoint_use = OF_timepoints)

# Calculate for all cutoff-compound-timepoint combinations
BR_ss <- purrr::pmap_dfr(param_grid, ~ make_calculations(dataset=BR, cutoff = ..1, compound = as.character(..2), timepoint_use = ..3))

BR_ss
```
```{r}
plot_cutoffs(dataset=BR_ss, 
             timepoint_use_variable=BR$timepoint, 
             tissue="Breath", 
             cpd="thc")
```
Compared to the Whole Blood and Oral Fluid, Breath seems to be only applicable for higher detection limits, which in this case is 10 ng/mL. And since we have already found that 5 ng/mL is already accurate in detecting with high specificity and sensitivity and we would like the lower dosage possible for detection, we can assess that the breath matrix is unnecessary for our study and has use for larger doses in the first three hours.

```{r}
# specify which calculations to make
cutoffs <- c(0.1, 5, 5.5, 6, 6.5)
compounds <- combined_user_long |> filter(fluid_type=="WB") |> filter(!is.na(value)) |> distinct(name) |> pull(name)
WB_timepoints <- c("pre-smoking", "0-60 min", "0-120 min", "181+ min")
WB <- combined |> filter(fluid_type=="WB")

# Specify all parameter combinations
param_grid <- expand.grid(
  cutoffs = cutoffs,
  compounds = compounds,
  timepoint_use = WB_timepoints)


# Calculate for all cutoff-compound-timepoint combinations
WB_new_ss <- purrr::pmap_dfr(param_grid, ~ make_calculations(dataset=WB, cutoff = ..1, compound = as.character(..2), timepoint_use = ..3))

WB_new_ss
```
```{r}
plot_cutoffs(dataset=WB_new_ss, 
             timepoint_use_variable=WB$timepoint, 
             tissue="Blood", 
             cpd="thc")
```
In context for cutoffs between 5 ng/mL -7 ng/mL, it has shown that higher detection limits do often become less sensitive to accuracy overtime, leading to a drop in likelihood to identify if THC has been recently used. The cutoffs in the middle have the most likelihood to be optimal in detecting THC in Blood that would be less sensitive to small traces of THC in the system that may not be causal for impairment, but is more likely to pick up a concentration of THC metabolites that is sustained in Blood for longer overtime.

## Results & Discussion 

The context of our question focuses on 
1. Which matrix–blood, oral fluid, or breath–is most effective for detecting recent THC use within a 3-hour window?
2. What is the optimal cutoff within this 3-hour timeframe for detecting THC in each matrix?
3. How does the treatment group (Placebo, 5.9% THC, or 13.4% THC) affect the detection of recent THC use in each matrix?





## Conclusion


[^1]: Pearson, J., et al. (2020). *Legal implications of THC per se limits.* Journal of Accident Analysis & Prevention. Retrieved from [Taylor & Francis Online](https://www.tandfonline.com/doi/full/10.1080/15389588.2020.1851685). 

[^2]: Huestis, M. A. (2007). *Human cannabinoid pharmacokinetics.* Chemistry & Biodiversity, 4(8), 1770-1804. [https://pmc.ncbi.nlm.nih.gov/articles/PMC2689518/). 

[^3]: Compton, R. (2017). *Marijuana-impaired driving: A report to Congress.* National Highway Traffic Safety Administration. Retrieved from [NGTSA](https://www.nhtsa.gov/sites/nhtsa.gov/files/documents/812440-marijuana-impaired-driving-report-to-congress.pdf).

[^4] Kosnett, M. J., Ma, M., Dooley, G., Wang, G. S., Friedman, K., Brown, T., Henthorn, T. K., & Brooks-Russell, A. (2023). Blood cannabinoid molar metabolite ratios are superior to blood THC as an indicator of recent cannabis smoking. Clinical Toxicology, 61(5), 355–362. https://doi.org/10.1080/15563650.2023.2214697 